---
title: "Data preparation for Blockbuster 2"
author: "Peter Curtis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Blockbuster2DataPrep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This code is used to prepare the raw PDS data (expected to be in csv files "./PDS/PDS_full_establishment.csv" "./PDS/PDS_full_building.csv", "./PDS/PDS_full_condition.csv") for use with the
[`Blockbuster2` Deterioration Model](https://github.com/DFE-Capital/Blockbuster-2).

The data is not supplied with this package.

The following code will create a single component-level
table containing deterioration rates, repair costs, and with 'empty' components
such as 'no decoration' removed.  This object uses the deterioration rates 
produced after the [QA by Roger Thompson](https://educationgovuk.sharepoint.com/:f:/r/sites/efacam/WorkplaceDocuments/04%20Condition%20and%20Cost%20Intelligence/14_Deterioration_Model/Roger%20Thompson%20Deterioration%20Rate%20QA?csf=1) and the repair costs from the [final PDSP Cost model](https://educationgovuk.sharepoint.com/sites/efacam/_layouts/15/DocIdRedir.aspx?ID=TXETKDDE3KEP-4-5171). 

```{r eval = FALSE}
create_PDS(single_table = TRUE, remove_elements = TRUE, add_rates = TRUE, add_costs = TRUE)
```

The following will produce an object with the 
data in a more efficient star schemaand without the deterioration rates or costs in readiness for the `Blockbuster2` version that enables input from an excel sheet that contains the relevant parameters.

```{r eval = FALSE}
create_PDS()
```
